Tree height and hydraulic traits shape growth responses across droughts in a temperate broadleaf forest

### 2020-09-06 manuscript/response_to_reviews.Rmd

The authors have satisfactorily answered most previous comments. For instance, I appreciated the discussion on the two indices Rt and Rtarima. I agree that results comparing them are pretty similar (Table S6 vs Table S7) and that authors can use Rt.

Yet, I still believe there are a couple of details that need revision:

1) I still think that adding the same analysis on the two other resilience indices (resilience and recovery) from Lloret et al. (2011) would increase the interest of the manuscript (as the authors acknowledge in line 476). Using DeSoto et al. 2020 as a justification is not enough. There are potential caveats in that paper and even if we accepted the general trends in growth resilience proposed in that study, studies like that of McGregor et al. apply at a different and more detailed scale, where those global relationships might not hold. This adds interest to the present ms, but for that same reason most readers will like to see a similar discussion on the other complementary growth resilience indices together in the same paper.

*([issue #131](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/131))*

@@

2) In Line 263 it is stated that both marginal and conditional R2 are reported, but then I could only find conditional R2s in Table S6, S7. Please, report also marginal R2 in those Tables to show how much variability it is explained by the fixed effects (i.e. the traits and independent variables in Table 3). This is important so we can judge how robust the tested covariates are. I can accept what authors argue in their response (‘Tree-ring data are inherently noisy, particularly in more mesic forests. In large part, this is due to strong neighborhood effects, along with variable allocation to ring growth around the circumference of the tree. These factors contribute to the low R2, but do not negate the significance of our findings’). However, still I think that, even if (or particularly because) we see many papers today with very low variability explained (see for instance some of the analyses in the previously mentioned DeSoto et al. 2020), we should demand a minimum level of explanatory power in models in ecophysiological papers. Otherwise we cannot assure that inferences are robust enough and relationships worth to be discussed.

*This is straightforward: need to report the marginal R2. Maybe also add some commentary on how a lot of noise doesn't negate significant signals.*

@@

Other comments:

- As mentioned on a previous comment, in the manuscript the authors sampled dead and live trees, but then I did not see any further mentioning to this later in the manuscript (lines 175-178 and Table 2). This would merit some discussion and further analysis, how many trees were died and how many trees were alive of those in Table 2?

*I think our argument is pretty strong, but we need to make some change to respond to this comment. Looking at resilience and recovery should help; presuming that they are high, we have a further argument that death was not caused by drought, and hence that status at coring doesn't matter.*

*All trees included in the analysis lived at least 17-18 years past the most recent drought. *

@@

- LMA is not shown in Figure S4, and in Table 2 there are no tests so we cannot judge whether there are differences among species for that leaf trait. Please, add also LMA to Figure S4.

*Ian, we should add both LMA and WD here.*

- Line 135: leave turgor ‘at’ lower ….

**Thanks for catching this; fixed. Sentence now reads, "We further test predictions that species with low $PLA_{dry}$ have higher drought resistance, and that species whose leaves lose turgor at lower water potentials (more negative $\pi_{tlp}$) have higher resistance."**

@@

- Line 272: please define dAIC. I think it should be better and more robust to stick in all cases to dAIC>2, there are only two covariates selected with <1 in Table S4.

**We now define AICc at first mention, which is in the paragraph above. The sentence now reads "We used Akaike information criterion with correction for small sample sizes (AICc) to assess model selection,..." In the sentence referenced here, we define $\Delta$AIC: "Trait variables were considered appropriate for inclusion in the main model if they had a consistent direction of response across all droughts and if their addition to the base model improved fit (at $\Delta$AICc $\ge$ 1.0, where $\Delta$AICc is the difference in AICc between models with and without the trait) in at least one drought year (Table S4)."**

**We have given careful thought to the trait variable selection criteria. It is important to note that selection of traits is not a test of their significance in the final model; it is simply a way of objectively reducing the number of candidate traits in these models. While we fully agree that $\Delta$AICc $\ge$ 2 is the appropriate criterion for assessing significance, it is not necessarily appropriate for candidate variable selection because variable significance changes when considered independently versus in the context of a full model. That is, the significance of variables--particularly those with marginal significance-- cannot be reliably assessed outside the context of a model that simultaneously accounts for other relevant variables. As a case in point, $\pi_{tlp}$ would be removed by a $\Delta$AICc $\ge$ 2 criterion for variable selection, but is included in the best model for all droughts combined, as well as the 1977 drought. While it's effects are marginal (i.e., included in some but not all top models), we feel that it is important to include it as a candidate variable both for the statistical considerations mentioned above, and because we expect that a lot of readers will be very interested to see a full assessment of this trait, which is currently of great interest to the research community.**

@@

- Line 312: please, refer to the corresponding Table or Supplement where this statement is supported.

**We have added this reference: "That is, drought year as a variable did not appear in any of the top models -- *i.e.*, models that were statistically indistinguishable ($\Delta$AICc<2) from the best model (see footnotes on Tables S6-S7)."**

@@

- Line 371-372: actually it is radial growth that is affected; whether that can be extrapolated to the tree level is something that needs to be demonstrated.

**This is true. Of course, tree biomass growth is almost always inferred from radial or diameter growth, sometimes in conjunction with height, using allometries. The issue of whether radial stem growth (or BAI, as examined here) actually translates directly to biomass growth goes far beyond the present study. To make sure this is clearly defined here, we've specified that we're considering stem growth in a couple key places:**

* **in the introduction, where we define resistance: "Here, we examine how tree height, microenvironment characteristics, and species' traits collectively shape drought resistance, defined as the ratio of annual stem growth in a drought year to that which would be expected in the absence of drought based on previous years' growth."**

* **in the sentence identified above: "These findings advance our knowledge of the factors that make trees vulnerable to stem growth declines during drought and, by extension, likely make them more vulnerable to mortality [@sapes_plant_2019]."**

Line 385 vs lines 308-309 and line 338: does that mean that there was a correlation between canopy position and positive Rt? Can it also be related to moisture sensitivity of different species, or to niche (microenvironment) partitioning among species? (as nicely discussed in lines 420-422). I see from lines 445-450 that sampled FAGR were small, which is coherent with the authors’ statement.

*(need to return here.)*

### 2020-09-06 manuscript/climate_sensitivity_draft.Rmd
worked on R3 comments

Here, we examine how tree height, microenvironment characteristics, and species' traits collectively shape drought resistance, defined as the ratio of annual growth in a drought year to that which would be expected in the absence of drought based on previous years' growth. 
We test a series of hypotheses and associated specific predictions (Table 1) based on the combination of tree-ring records from the three strongest droughts over a 60-year period (1950 - 2009), species trait measurements, and census and microenvironmental data from a large forest dynamics plot in Virginia, USA. 
First, we focus on how tree size, alone and in its interaction with microenvironmental gradients, influences drought resistance. 
We examine the contemporary relationship between tree height and microenvironment, including growing season meteorological conditions and crown exposure. 
We then test whether, consistent with most forests globally, larger-diameter, taller trees tend to have lower drought resistance in this forest, which is in a region (eastern North America) represented by only two studies in the global review of [@bennett_larger_2015].
We also test for an influence of potential access to available soil water, which should be greater for larger trees in dry but not in perpetually wet microsites. 
Finally, we focus on the role of species' traits, testing the hypothesis that species’ traits-–particularly leaf drought tolerance traits-–predict drought resistance. 
We test predictions that drought resistance is higher in ring-porous than semi-ring and diffuse-porous species and that it is correlated with wood density--either positively [@greenwood_tree_2017] or negatively [@hoffmann_hydraulic_2011] and positively correlated with $LMA$.
We further test predictions that species with low $PLA_{dry}$ have higher drought resistance, and that species whose leaves lose turgor lower water potentials (more negative $\pi_{tlp}$) have higher resistance.

@@

Analyses focused on testing the predictions presented in Table 1 with $Rt$ as the response variable, and then repeated using $Rt_{ARIMA}$ as the response variable. 
Models were run for all drought years combined and for each drought year individually. 
The general statistical model for hypothesis testing was a mixed effects model, implemented in the lme4 package in R [@R-lme4].
In the multi-year model, we included a random effect of tree nested within species and a fixed effect of drought year to represent the combined effects of differences in drought characteristics.
Individual year models included a random effect of species. 
All models included fixed effects of independent variables of interest (Tables 1,3) as specified below.
All variables across all best models had variance inflation factors <1.2 (1 +/- 0.019). 
We used AICc to assess model selection, and conditional/marginal R-squared to assess model fit as implemented in the AICcmodavg package in R [@R-AICcmodavg]. 
AICc refers to a corrected version of AICc, and is best suited for small data sizes [see @brewer_relative_2016].

To avoid over-fitting models with five species traits (Table 3) across only 12 species, we did not include all traits as fixed effects in a single linear mixed model, but rather conducted individual tests of each species trait to determine the relative importance and appropriateness for inclusion in the main model. 
These tests followed the model structure specified above, then added $ln[H]$ and $ln[TWI]$ to create a base model against which we tested traits.
Trait variables were considered appropriate for inclusion in the main model if they had a consistent direction of response across all droughts and if their addition to the base model improved fit (at $\Delta$AICc $\ge$ 1.0) in at least one drought year (Table S4). 
We note that we did not use the $\Delta$AICc $\ge$ 1.0 criterion as a test of significance, but rather of whether the variable had enough influence to be considered as a *candidate* variable in full models.

@@

In the context of the multivariate model, $Rt$ did not vary across drought years. That is, drought year as a variable did not appear in any of the top models -- *i.e.*, models that were statistically indistinguishable ($\Delta$AICc<2) from the best model.

@@

In contrast, $PLA_{dry}$, and $\pi_{tlp}$ - the traits that qualified for inclusion in the full model (Table S4) - were both negatively correlated to drought resistance (Figs. **4, S6**; Tables 1,S4-S7). 
$PLA_{dry}$ had a significant influence, with negative coefficient, in full models for the three droughts combined and for the 1966 drought individually (Fig. **4**; Tables S6-S7). 
For 1977 and 1999, it was included with a negative coefficient in some of the top models (Tables S6-S7).
$\pi_{tlp}$ was included with a negative coefficient in the best model for both all droughts combined and for the 1977 drought individually (Fig. **4**; Table S6). 
It was also included in some of the top models for 1999 (Tables S6-S7).

@@

Tree height, microenvironment, and leaf drought tolerance traits shaped tree growth responses across three droughts at our study site (Table 1, Fig. **4**). The greater susceptibility of larger trees to drought, similar to forests worldwide [@bennett_larger_2015], was driven primarily by their height [@stovall_tree_2019]. Taller height was likely a liability in itself, and was also associated with greater exposure to conditions that would promote water loss and heat damage during drought (Fig. **2**). There was no evidence that greater availability of, or access to, soil water availability increased drought resistance; in contrast, trees in wetter topographic positions had lower $Rt$ [@zuleta_drought-induced_2017; @stovall_tree_2019], and the larger potential rooting volume of large trees provided no advantage in the drier microenvironments. The negative effect of height on $Rt$ held after accounting for species' traits, which is consistent with recent work finding height had a stronger influence on mortality risk than forest type during drought [@stovall_reply_2020]. Drought resistance was not consistently linked to species' $LMA$, wood density, or xylem type (ring- vs. diffuse porous), but was negatively correlated with leaf drought tolerance traits ($PLA_{dry}$, $\pi_{tlp}$). This is the first study to our knowledge linking $PLA_{dry}$ and $\pi_{tlp}$ to growth reduction during drought. The directions of these responses were consistent across droughts (Table S6), supporting the premise that they were driven by fundamental physiological mechanisms. However, the strengths of each predictor varied across droughts (Fig. **4**; Tables S6-S7), indicating that drought characteristics interact with tree size, microenvironment, and traits to shape which individuals are most affected. These findings advance our knowledge of the factors that make trees vulnerable to growth declines during drought and, by extension, likely make them more vulnerable to mortality [@sapes_plant_2019].

>>>>

Here, we examine how tree height, microenvironment characteristics, and species' traits collectively shape drought resistance, defined as the ratio of annual stem growth in a drought year to that which would be expected in the absence of drought based on previous years' growth. 
We test a series of hypotheses and associated specific predictions (Table 1) based on the combination of tree-ring records from the three strongest droughts over a 60-year period (1950 - 2009), species trait measurements, and census and microenvironmental data from a large forest dynamics plot in Virginia, USA. 
First, we focus on how tree size, alone and in its interaction with microenvironmental gradients, influences drought resistance. 
We examine the contemporary relationship between tree height and microenvironment, including growing season meteorological conditions and crown exposure. 
We then test whether, consistent with most forests globally, larger-diameter, taller trees tend to have lower drought resistance in this forest, which is in a region (eastern North America) represented by only two studies in the global review of [@bennett_larger_2015].
We also test for an influence of potential access to available soil water, which should be greater for larger trees in dry but not in perpetually wet microsites. 
Finally, we focus on the role of species' traits, testing the hypothesis that species’ traits-–particularly leaf drought tolerance traits-–predict drought resistance. 
We test predictions that drought resistance is higher in ring-porous than semi-ring and diffuse-porous species and that it is correlated with wood density--either positively [@greenwood_tree_2017] or negatively [@hoffmann_hydraulic_2011] and positively correlated with $LMA$.
We further test predictions that species with low $PLA_{dry}$ have higher drought resistance, and that species whose leaves lose turgor at lower water potentials (more negative $\pi_{tlp}$) have higher resistance.

@@

Analyses focused on testing the predictions presented in Table 1 with $Rt$ as the response variable, and then repeated using $Rt_{ARIMA}$ as the response variable. 
Models were run for all drought years combined and for each drought year individually. 
The general statistical model for hypothesis testing was a mixed effects model, implemented in the lme4 package in R [@R-lme4].
In the multi-year model, we included a random effect of tree nested within species and a fixed effect of drought year to represent the combined effects of differences in drought characteristics.
Individual year models included a random effect of species. 
All models included fixed effects of independent variables of interest (Tables 1,3) as specified below.
All variables across all best models had variance inflation factors <1.2 (1 +/- 0.019). 
We used Akaike information criterion with correction for small sample sizes (AICc) to assess model selection, and conditional/marginal R-squared to assess model fit as implemented in the AICcmodavg package in R [@R-AICcmodavg]. 
AICc refers to a corrected version of AICc, and is best suited for small data sizes [see @brewer_relative_2016].

To avoid over-fitting models with five species traits (Table 3) across only 12 species, we did not include all traits as fixed effects in a single linear mixed model, but rather conducted individual tests of each species trait to determine the relative importance and appropriateness for inclusion in the main model. 
These tests followed the model structure specified above, then added $ln[H]$ and $ln[TWI]$ to create a base model against which we tested traits.
Trait variables were considered appropriate for inclusion in the main model if they had a consistent direction of response across all droughts and if their addition to the base model improved fit (at $\Delta$AICc $\ge$ 1.0, where $\Delta$AICc is the difference in AICc between models with and without the trait) in at least one drought year (Table S4). 
We note that we did not use the $\Delta$AICc $\ge$ 1.0 criterion as a test of significance, but rather of whether the variable had enough influence to be considered as a *candidate* variable in full models.

@@

In the context of the multivariate model, $Rt$ did not vary across drought years. That is, drought year as a variable did not appear in any of the top models -- *i.e.*, models that were statistically indistinguishable ($\Delta$AICc<2) from the best model (see footnotes on Tables S6-S7).

@@

In contrast, $PLA_{dry}$, and $\pi_{tlp}$ - the traits that qualified for inclusion in the full model (Table S4) - were both negatively correlated to drought resistance (Figs. **4, S6**; Tables 1,S4-S7). 
$PLA_{dry}$ had a significant influence, with negative coefficient, in full models for the three droughts combined and for the 1966 drought individually (Fig. **4**; Tables S6-S7). 
For 1977 and 1999, it was included with a negative coefficient in some of the top models (Tables S6-S7).
$\pi_{tlp}$ was included with a negative coefficient in the best model for both all droughts combined and for the 1977 drought individually (Fig. **4**; Table S6), although its influence was not significant at $\Delta$AICc<2. 
It was also included in some of the top models for 1999 (Tables S6-S7).

@@

Tree height, microenvironment, and leaf drought tolerance traits shaped tree growth responses across three droughts at our study site (Table 1, Fig. **4**). The greater susceptibility of larger trees to drought, similar to forests worldwide [@bennett_larger_2015], was driven primarily by their height [@stovall_tree_2019]. Taller height was likely a liability in itself, and was also associated with greater exposure to conditions that would promote water loss and heat damage during drought (Fig. **2**). There was no evidence that greater availability of, or access to, soil water availability increased drought resistance; in contrast, trees in wetter topographic positions had lower $Rt$ [@zuleta_drought-induced_2017; @stovall_tree_2019], and the larger potential rooting volume of large trees provided no advantage in the drier microenvironments. The negative effect of height on $Rt$ held after accounting for species' traits, which is consistent with recent work finding height had a stronger influence on mortality risk than forest type during drought [@stovall_reply_2020]. Drought resistance was not consistently linked to species' $LMA$, wood density, or xylem type (ring- vs. diffuse porous), but was negatively correlated with leaf drought tolerance traits ($PLA_{dry}$, $\pi_{tlp}$). This is the first study to our knowledge linking $PLA_{dry}$ and $\pi_{tlp}$ to growth reduction during drought. The directions of these responses were consistent across droughts (Table S6), supporting the premise that they were driven by fundamental physiological mechanisms. However, the strengths of each predictor varied across droughts (Fig. **4**; Tables S6-S7), indicating that drought characteristics interact with tree size, microenvironment, and traits to shape which individuals are most affected. These findings advance our knowledge of the factors that make trees vulnerable to stem growth declines during drought and, by extension, likely make them more vulnerable to mortality [@sapes_plant_2019].
